---
title: "Preliminary Plots 3"
output: html_document
date: "2024-06-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load, include=FALSE}
library(tidyverse)
lw_data <- readRDS('Desktop/GMRI/data/bottomTrawlSurvey_indLengthWeight.rds')
```

## Splitting by Season
<br>
This analysis groups fish by season as well as decade to determine significance.
<br>
## Atlantic Cod
<br>
Original Cod data:
<br>
```{r cod plots, echo=FALSE}
cod <- lw_data %>%
  filter(COMNAME == 'ATLANTIC COD') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2020 ~ "2020s",
                            YEAR >= 2010 & YEAR < 2020 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

cod_orig <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  labs(title = "Atlantic Cod: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(cod_orig)

cod_color <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  facet_wrap(~decade)+
  labs(title = "Atlantic Cod: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(cod_color)

cod_split <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point()+
  facet_wrap(~SEASON)+
  labs(title = "Atlantic Cod: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(cod_split)
```
<br>
Transformed Cod data:
<br>
```{r cod log plots, echo=FALSE}
cod_log <- lw_data %>%
  filter(COMNAME == 'ATLANTIC COD') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2020 ~ "2020s",
                            YEAR >= 2010 & YEAR < 2020 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

cod_log_orig <- cod_log %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  labs(title = "Atlantic Cod: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(cod_log_orig)

cod_log_color <- cod_log %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  facet_wrap(~decade)+
  labs(title = "Atlantic Cod: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(cod_log_color)

cod_log_split <- cod_log %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point()+
  facet_wrap(~SEASON)+
  labs(title = "Atlantic Cod: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(cod_log_split)
```
<br>

Cod table by decade and season and linear model:

```{r cod model, echo=FALSE}
cod.b.by.dec <- matrix(nrow=4, ncol=4, byrow=TRUE)
rownames(cod.b.by.dec) <- 
  c('SUMMER', 'FALL', 'WINTER', 'SPRING')
colnames(cod.b.by.dec) <- 
  c('1990s', '2000s', '2010s', '2020s')

cod.a.by.dec <- matrix(nrow=4, ncol=4, byrow=TRUE)
rownames(cod.a.by.dec) <- 
  c('SUMMER', 'FALL', 'WINTER', 'SPRING')
colnames(cod.a.by.dec) <- 
  c('1990s', '2000s', '2010s', '2020s')

decades = c("1990s", "2000s", "2010s", "2020s")
seasons = c("SUMMER", "SPRING", "AUTUMN", "WINTER")
for (dec in decades) {
  for (szn in seasons) {
    data <- cod %>%
      filter(decade == dec & SEASON == szn)
    if (nrow(data) > 5) {
    lin.mod <- lm(log(INDWT) ~ log(LENGTH), data)
    cod.b.by.dec[szn, dec] <- summary(lin.mod)$coefficients[2,1]
    cod.a.by.dec[szn, dec] <- summary(lin.mod)$coefficients[1,1]
    }
  }
}
print(cod.b.by.dec)

bcod_df <- as.data.frame(cod.b.by.dec["SPRING", , drop = FALSE])
bcod_long <- pivot_longer(bcod_df, 
                       cols = c("1990s", "2000s", "2010s", "2020s"), 
                       names_to = "Decade", 
                       values_to = "Slope")
ggplot(bcod_long, aes(x = Decade, y = Slope, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Slope of Relationship Between log(Length) and log(Weight) for Cod in Spring",
       x = "Decade",
       y = "Slope (b)") +
  theme_minimal()

cod_model <- lm(log(cod$INDWT) ~ log(cod$LENGTH)*cod$decade + log(cod$LENGTH)*cod$SEASON)
summary(cod_model)
```

<br>

## American Plaice

```{r plaice plots, echo=FALSE}
plaice <- lw_data %>%
  filter(COMNAME == 'AMERICAN PLAICE') %>%
  filter(INDWT != 0 & INDWT < 4) %>%
  mutate(decade = case_when(YEAR >= 2020 ~ "2020s",
                            YEAR >= 2010 & YEAR < 2020 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

plaice_orig <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  labs(title = "Atlantic plaice: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(plaice_orig)

plaice_color <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  facet_wrap(~decade)+
  labs(title = "Atlantic plaice: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(plaice_color)

plaice_split <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point()+
  facet_wrap(~SEASON)+
  labs(title = "Atlantic plaice: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(plaice_split)
```
<br>
Transformed American plaice data:
<br>
```{r plaice log plots, echo=FALSE}
Plaice_log <- lw_data %>%
  filter(COMNAME == 'AMERICAN PLAICE') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2020 ~ "2020s",
                            YEAR >= 2010 & YEAR < 2020 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

Plaice_log_orig <- Plaice_log %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  labs(title = "American Plaice: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(Plaice_log_orig)

Plaice_log_color <- Plaice_log %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  facet_wrap(~decade)+
  labs(title = "American Plaice: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(Plaice_log_color)

Plaice_log_split <- Plaice_log %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point()+
  facet_wrap(~SEASON)+
  labs(title = "American Plaice: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(Plaice_log_split)
```
<br>

American plaice table by decade and season and linear model:

```{r plaice model, echo=FALSE}
plaice.b.by.dec <- matrix(nrow=3, ncol=4, byrow=TRUE)
rownames(plaice.b.by.dec) <- 
  c('SUMMER', 'WINTER', 'SPRING')
colnames(plaice.b.by.dec) <- 
  c('1990s', '2000s', '2010s', '2020s')

plaice.a.by.dec <- matrix(nrow=3, ncol=4, byrow=TRUE)
rownames(plaice.a.by.dec) <- 
  c('SUMMER', 'WINTER', 'SPRING')
colnames(plaice.a.by.dec) <- 
  c('1990s', '2000s', '2010s', '2020s')

decades = c("1990s", "2000s", "2010s", "2020s")
seasons = c("SUMMER", "SPRING", "AUTUMN", "WINTER")
for (dec in decades) {
  for (szn in seasons) {
    data <- plaice %>%
      filter(decade == dec & SEASON == szn)
    if (nrow(data) > 5) {
    lin.mod <- lm(log(INDWT) ~ log(LENGTH), data)
    plaice.b.by.dec[szn, dec] <- summary(lin.mod)$coefficients[2,1]
    plaice.a.by.dec[szn, dec] <- summary(lin.mod)$coefficients[1,1]
    }
  }
}
print(plaice.b.by.dec)

plaice_model <- lm(log(plaice$INDWT) ~ log(plaice$LENGTH)*plaice$decade + log(plaice$LENGTH)*plaice$SEASON)
summary(plaice_model)
```
<br>
## Atlantic Herring
<br>
Original Atlantic herring data plots:
<br>
```{r herring plots, echo=FALSE}
herring <- lw_data %>%
  filter(COMNAME == 'ATLANTIC HERRING') %>%
  filter(INDWT > 0 & INDWT < 0.5) %>%
  mutate(decade = case_when(YEAR >= 2020 ~ "2020s",
                            YEAR >= 2010 & YEAR < 2020 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

herring_orig <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  labs(title = "Atlantic herring: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(herring_orig)

herring_color <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  facet_wrap(~decade)+
  labs(title = "Atlantic herring: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(herring_color)

herring_split <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point()+
  facet_wrap(~SEASON)+
  labs(title = "Atlantic herring: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(herring_split)
```
<br>
Transformed Atlantic herring data:
<br>
```{r herring log plots, echo=FALSE}
herring_log_orig <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  labs(title = "Atlantic herring: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(herring_log_orig)

herring_log_color <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  facet_wrap(~decade)+
  labs(title = "Atlantic herring: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(herring_log_color)

herring_log_split <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point()+
  facet_wrap(~SEASON)+
  labs(title = "Atlantic herring: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(herring_log_split)
```
<br>

Atlantic herring table by decade and season and linear model:

```{r herring model, echo=FALSE}
herring.b.by.dec <- matrix(nrow=4, ncol=4, byrow=TRUE)
rownames(herring.b.by.dec) <- 
  c('SUMMER', 'FALL', 'WINTER', 'SPRING')
colnames(herring.b.by.dec) <- 
  c('1990s', '2000s', '2010s', '2020s')

herring.a.by.dec <- matrix(nrow=4, ncol=4, byrow=TRUE)
rownames(herring.a.by.dec) <- 
  c('SUMMER', 'FALL', 'WINTER', 'SPRING')
colnames(herring.a.by.dec) <- 
  c('1990s', '2000s', '2010s', '2020s')

decades = c("1990s", "2000s", "2010s", "2020s")
seasons = c("SUMMER", "SPRING", "AUTUMN", "WINTER")
for (dec in decades) {
  for (szn in seasons) {
    data <- herring %>%
      filter(decade == dec & SEASON == szn)
    if (nrow(data) > 5) {
    lin.mod <- lm(log(INDWT) ~ log(LENGTH), data)
    herring.b.by.dec[szn, dec] <- summary(lin.mod)$coefficients[2,1]
    herring.a.by.dec[szn, dec] <- summary(lin.mod)$coefficients[1,1]
    }
  }
}
print(herring.b.by.dec)

herring_model <- lm(log(herring$INDWT) ~ log(herring$LENGTH)*herring$decade + log(herring$LENGTH)*herring$SEASON)
summary(herring_model)
```

<br>

## Spiny Dogfish

```{r dogfish plots, echo=FALSE}
dogfish <- lw_data %>%
  filter(COMNAME == 'SPINY DOGFISH') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2020 ~ "2020s",
                            YEAR >= 2010 & YEAR < 2020 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

dogfish_orig <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  labs(title = "Atlantic dogfish: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(dogfish_orig)

dogfish_color <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  facet_wrap(~decade)+
  labs(title = "Atlantic dogfish: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(dogfish_color)

dogfish_split <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point()+
  facet_wrap(~SEASON)+
  labs(title = "Atlantic dogfish: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(dogfish_split)
```
<br>
Transformed dogfish data:
<br>
```{r dogfish log plots, echo=FALSE}
dogfish_log_orig <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  labs(title = "Spiny dogfish: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(dogfish_log_orig)

dogfish_log_color <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=SEASON))+
  facet_wrap(~decade)+
  labs(title = "Spiny dogfish: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(dogfish_log_color)

dogfish_log_split <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point()+
  facet_wrap(~SEASON)+
  labs(title = "Spiny dogfish: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(dogfish_log_split)
```
<br>

Spiny dogfish table by decade and season and linear model:

```{r dogfish model, echo=FALSE}
dogfish.b.by.dec <- matrix(nrow=4, ncol=4, byrow=TRUE)
rownames(dogfish.b.by.dec) <- 
  c('SUMMER', 'FALL', 'WINTER', 'SPRING')
colnames(dogfish.b.by.dec) <- 
  c('1990s', '2000s', '2010s', '2020s')

dogfish.a.by.dec <- matrix(nrow=4, ncol=4, byrow=TRUE)
rownames(dogfish.a.by.dec) <- 
  c('SUMMER', 'FALL', 'WINTER', 'SPRING')
colnames(dogfish.a.by.dec) <- 
  c('1990s', '2000s', '2010s', '2020s')

decades = c("1990s", "2000s", "2010s", "2020s")
seasons = c("SUMMER", "SPRING", "AUTUMN", "WINTER")
for (dec in decades) {
  for (szn in seasons) {
    data <- dogfish %>%
      filter(decade == dec & SEASON == szn)
    if (nrow(data) > 5) {
    lin.mod <- lm(log(INDWT) ~ log(LENGTH), data)
    dogfish.b.by.dec[szn, dec] <- summary(lin.mod)$coefficients[2,1]
    dogfish.a.by.dec[szn, dec] <- summary(lin.mod)$coefficients[1,1]
    }
  }
}
print(dogfish.b.by.dec)

dogfish_model <- lm(log(dogfish$INDWT) ~ log(dogfish$LENGTH)*dogfish$decade + log(dogfish$LENGTH)*dogfish$SEASON)
summary(dogfish_model)
```