---
title: "Preliminary Plots & Comments"
output: html_document
date: "2024-06-20"
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading in, echo=FALSE}
library(tidyverse)
lw_data <- readRDS('data/bottomTrawlSurvey_indLengthWeight.rds')
lw_coeffs <- readRDS('data/lw_coeffs.rds')

# create tables to hold parameter values
b.by.decade <- matrix(nrow=4, ncol=3, byrow=TRUE)
rownames(b.by.decade) <- 
  c('Cod', 'American Plaice', 'Atlantic Herring', 'Spiny Dogfish')
colnames(b.by.decade) <- 
  c('1990s', '2000s', '2010s')

a.by.decade <- matrix(nrow=4, ncol=3, byrow=TRUE)
rownames(a.by.decade) <- 
  c('Cod', 'American Plaice', 'Atlantic Herring', 'Spiny Dogfish')
colnames(a.by.decade) <- 
  c('1990s', '2000s', '2010s')

cod <- lw_data %>%
  filter(COMNAME == 'ATLANTIC COD') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2010 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))
```

## Cod Plots & Regression

Original cod data, length vs. weight: a clear exponential pattern can be seen.

```{r original cod data plot, echo=FALSE}
### ORIGINAL DATA PLOT
cod_orig <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point() +
  labs(title = "Atlantic Cod: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(cod_orig)
```
<br>

After a log-transformation is applied, the data show a clear linear pattern:

```{r log transformed cod data, echo=FALSE}
cod_log <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point() +
  labs(title = "Atlantic Cod: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(cod_log)
```
<br>

Original data, color-coded by decade: this plot hints that fish are not reaching the same weights in the 2010s that they did in the two decades prior, yet the general trend appears to be about the same.

```{r original cod data by decade (color), echo=FALSE}
cod_dec.by.color.orig <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point(aes(color = decade))+
  labs(title = "Atlantic Cod: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(cod_dec.by.color.orig)
```
<br>

Transformed data, color-coded by decade: this plot makes it a little harder to see a difference by decade, but now the data is linear.

```{r transformed cod data by decade (colored), echo=FALSE}
cod_dec.by.color.log <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point(aes(color = decade))+
  labs(title = "Atlantic Cod: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(cod_dec.by.color.log)
```
<br>

Original data, split by decade: this plot may be most effective at showing the difference by decade. It does appear that cod in the 2000s may have had the best condition of the three decades, but it appears that the 2010s cod have a worse condition than cod in earlier decades. 

```{r original cod data by decade (split), echo=FALSE}
cod_dec.split.orig <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point()+
  facet_wrap(~decade)+
  labs(title = "Atlantic Cod: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(cod_dec.split.orig)
```
<br>

Transformed data, split by decade: again, harder to see a difference in the log-transformed plots, but we see that the 2000s decade has slightly more variability in weights here.

```{r transformed cod data by decade (split), echo=FALSE}
cod_dec.split.log <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point()+
  facet_wrap(~decade)+
  labs(title = "Atlantic Cod: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(cod_dec.split.log)
```
<br>

Cod regression model: we see some significance, but not with the 2010s decade.

```{r cod regression, echo=FALSE}
cod_90s <- cod %>%
  filter(decade == '1990s')
cod_90s_lm <- lm(log(INDWT) ~ log(LENGTH), cod_90s)
a.by.decade['Cod', '1990s'] <- summary(cod_90s_lm)$coefficients[1,1]
b.by.decade['Cod', '1990s'] <- summary(cod_90s_lm)$coefficients[2,1]

cod_00s <- cod %>%
  filter(decade == '2000s')
cod_00s_lm <- lm(log(INDWT) ~ log(LENGTH), cod_00s)
a.by.decade['Cod', '2000s'] <- summary(cod_00s_lm)$coefficients[1,1]
b.by.decade['Cod', '2000s'] <- summary(cod_00s_lm)$coefficients[2,1]

cod_10s <- cod %>%
  filter(decade == '2010s')
cod_10s_lm <- lm(log(INDWT) ~ log(LENGTH), cod_10s)
a.by.decade['Cod', '2010s'] <- summary(cod_10s_lm)$coefficients[1,1]
b.by.decade['Cod', '2010s'] <- summary(cod_10s_lm)$coefficients[2,1]

# actual model!
cod_model <- lm(log(INDWT) ~ log(LENGTH)*decade, cod)
summary(cod_model)
```
<br>
Cod predictions plot, based on the model: there appears to be a minimal difference in the slopes.

```{r cod predictions, echo=FALSE}
# generating predictions for the model
new_cod = expand.grid(LENGTH = seq(min(cod$LENGTH), max(cod$LENGTH), length.out = 100), decade = c("1990s", "2000s", "2010s")) 
new_cod$predicted <- exp(predict(cod_model, newdata = new_cod, type = "response")) 
new_cod$lower <- exp(predict(cod_model, newdata = new_cod, type = "response", interval = 
                           "confidence")[,3])
new_cod$upper <- exp(predict(cod_model, newdata = new_cod, type = "response", interval = 
                           "confidence")[,2])
ggplot(new_cod, aes(x = LENGTH, y = predicted))+
  geom_line(aes(color = decade))+
  geom_ribbon(aes(group = decade, ymax = upper, ymin = lower), alpha = 0.1)+ 
  scale_x_log10()+
  scale_y_log10()+ 
  theme_classic()+
  labs(title = "Cod Predicted Weights by Length",
       x = "Length",
       y = "Individual Weight")
```

## American Plaice Plots & Regression

As large cod are frequently fished and thus likely underrepresented in the trawl survey data, we turn to analyze another species who follows the periodic life history strategy whose data may be more reliable.

Original American plaice data, length vs. weight: the slope of this relationship appears to be far less than that of the cod species, but the same general exponential trend remains. There appears to be an outlier at length 10, weight 5: based on the rest of the data, this observation may be an error in the data.

```{r original plaice plot, echo=FALSE}
plaice <- lw_data %>%
  filter(COMNAME == 'AMERICAN PLAICE') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2010 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

### ORIGINAL DATA PLOT
plaice_orig <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point() +
  labs(title = "American Plaice: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(plaice_orig)
```
<br>

After a log-transformation is applied, the data appears to follow a linear trend.

```{r transformed plaice data plot, echo=FALSE}
plaice_log <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point() +
  labs(title = "American Plaice: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(plaice_log)
```
<br>

Original data, color-coded by decade: it appears that the weight distribution of plaice in the 2010s differs slightly from prior decades, but again follows the same general exponential trend as the 1990s and 2000s plaice. The outlier observation is from the 2000s. 

```{r plaice data by decade (color), echo=FALSE}
plaice_dec.by.color.orig <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point(aes(color = decade))+
  labs(title = "American Plaice: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(plaice_dec.by.color.orig)

```

<br>

Transformed data, color-coded by decade: even with the data transformed, 2010s plaice appear to be weighing less at longer lengths than plaice from the 1990s and 2000s. 

```{r transformed plaice data (color), echo=FALSE}
plaice_dec.by.color.log <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point(aes(color = decade))+
  labs(title = "American Plaice: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(plaice_dec.by.color.log)
```
<br>

Original data, split by decade: the slope of this relationship appears to be decreasing by decade.

```{r original plaice by decade (color), echo=FALSE}
plaice_dec.split.orig <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point()+
  facet_wrap(~decade)+
  labs(title = "American Plaice: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(plaice_dec.split.orig)
```
<br>

Transformed data, split by decade: the slopes appear to be slightly different here, but again the relationship is harder to see in this log-transformed plot.

```{r transformed plaice by decade (color), echo=FALSE}
plaice_dec.split.log <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point()+
  facet_wrap(~decade)+
  labs(title = "American Plaice: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(plaice_dec.split.log)
```
<br>

American plaice regression model: we see that all terms in the model are very significant. Log(weight) decreases significantly in the 2000s and 2010s, but interestingly enough it appeared to decrease more in the 2000s than the 2010s. Either way, a significant change decade-to-decade exists.

```{r plaice regression, echo=FALSE}
plaice_90s <- plaice %>%
  filter(decade == '1990s')
plaice_90s_lm <- lm(log(INDWT) ~ log(LENGTH), plaice_90s)
a.by.decade['American Plaice', '1990s'] <- summary(plaice_90s_lm)$coefficients[1,1]
b.by.decade['American Plaice', '1990s'] <- summary(plaice_90s_lm)$coefficients[2,1]

plaice_00s <- plaice %>%
  filter(decade == '2000s')
plaice_00s_lm <- lm(log(INDWT) ~ log(LENGTH), plaice_00s)
a.by.decade['American Plaice', '2000s'] <- summary(plaice_00s_lm)$coefficients[1,1]
b.by.decade['American Plaice', '2000s'] <- summary(plaice_00s_lm)$coefficients[2,1]

plaice_10s <- plaice %>%
  filter(decade == '2010s')
plaice_10s_lm <- lm(log(INDWT) ~ log(LENGTH), plaice_10s)
a.by.decade['American Plaice', '2010s'] <- summary(plaice_10s_lm)$coefficients[1,1]
b.by.decade['American Plaice', '2010s'] <- summary(plaice_10s_lm)$coefficients[2,1]

# actual model!
plaice_model <- lm(log(INDWT) ~ log(LENGTH)*decade, plaice)
summary(plaice_model)
```
<br>
American plaice predictions plot, based on the model: a slight difference in the decadal predictions in this plot can be seen, but again they are minor.

```{r plaice predictions plot, echo=FALSE}
# generating predictions for the model
new_plaice = expand.grid(LENGTH = seq(min(plaice$LENGTH), max(plaice$LENGTH), length.out = 100), decade = c("1990s", "2000s", "2010s")) 
new_plaice$predicted <- exp(predict(plaice_model, newdata = new_plaice, type = "response")) 
new_plaice$lower <- exp(predict(plaice_model, newdata = new_plaice, type = "response", interval = 
                               "confidence")[,3])
new_plaice$upper <- exp(predict(plaice_model, newdata = new_plaice, type = "response", interval = 
                               "confidence")[,2])
ggplot(new_plaice, aes(x = LENGTH, y = predicted))+
  geom_line(aes(color = decade))+
  geom_ribbon(aes(group = decade, ymax = upper, ymin = lower), alpha = 0.1)+ 
  scale_x_log10()+
  scale_y_log10()+ 
  theme_classic()+
  labs(title = "American Plaice Predicted Weights by Length",
       x = "Length",
       y = "Individual Weight")
```


## Atlantic Herring

We now examine an opportunistic species, the Atlantic herring.

Original Atlantic herring data, length vs. weight: the same general exponential trend between length and weight can be seen here, but the variance is far different than it was for both of the periodic fish analyzed above: the data are much more spread out.

```{r original herring data plot, echo=FALSE}
herring <- lw_data %>%
  filter(COMNAME == 'ATLANTIC HERRING') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2010 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

### ORIGINAL DATA
herring_orig <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point() +
  labs(title = "Atlantic Herring: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()+
  ylim(0, 0.75) # had to add - some HEAVY outliers
print(herring_orig)
```
<br> 
Also, there are some HEAVY outliers. Here's what the plot looked like before I limited the y-axis bounds:

```{r original herring w outliers, echo=FALSE}
herring_2 <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point() +
  labs(title = "Atlantic Herring: Length vs Weight (with outliers)",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(herring_2)
```

I have to imagine those are errors in the data...

<br>

After a log-transformation is applied, the data appear more linear. This plot is limited as well, so some outliers may be left off.

```{r transformed herring data plot, echo=FALSE}
### TRANSFORMED DATA
herring_log <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point() +
  labs(title = "Atlantic Herring: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(herring_log)
```
<br>

Original data, color-coded by decade: the variation in weight appears to differ by decade based on this plot: 2010s data are more concentrated.

```{r original herring data by decade (color), echo=FALSE}
herring_dec.by.color.orig <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point(aes(color = decade))+
  labs(title = "Atlantic Herring Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()+
  ylim(0, 0.75)
print(herring_dec.by.color.orig)
```
<br>

Transformed data, color-coded by decade: a slight decadal difference can be seen in this plot. It may be one of the clearer log-transformed plots produced in this analysis.

```{r transformed herring by decade (color), echo=FALSE}
herring_dec.by.color.log <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point(aes(color = decade))+
  labs(title = "Atlantic Herring: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(herring_dec.by.color.log)
```

<br>

Original data, split by decade: again, we see a more concentrated cluster in the 2010s compared to earlier decades.

```{r original herring data by decade (split), echo=FALSE}
### ORIGINAL DATA - SPLIT BY DECADE
herring_dec.split.orig <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point()+
  facet_wrap(~decade)+
  labs(title = "Atlantic Herring: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()+
  ylim(0, 0.75)
print(herring_dec.split.orig)
```
<br>

Transformed data, split by decade: here we see some outliers in the 2010s come into the picture as we transition to the log-scale.

```{r transformed herring data by decade (split), echo=FALSE}
herring_dec.split.log <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point()+
  facet_wrap(~decade)+
  labs(title = "Atlantic Herring: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(herring_dec.split.log)
```
<br>

Herring regression model: again we see that decade is significant in the model. It appears the fish are weighing significantly more in the 2010s than in the 1990s, but this is likely due to the outliers discussed above - probably should remove those from the dataset and rerun this regression, and these are almost certainly erroneous points.

```{r herring model, echo=FALSE}
# actual statistics stuff again! - saving parameters
herring_90s <- herring %>%
  filter(decade == '1990s')
herring_90s_lm <- lm(log(INDWT) ~ log(LENGTH), herring_90s)
a.by.decade['Atlantic Herring', '1990s'] <- summary(herring_90s_lm)$coefficients[1,1]
b.by.decade['Atlantic Herring', '1990s'] <- summary(herring_90s_lm)$coefficients[2,1]

herring_00s <- herring %>%
  filter(decade == '2000s')
herring_00s_lm <- lm(log(INDWT) ~ log(LENGTH), herring_00s)
a.by.decade['Atlantic Herring', '2000s'] <- summary(herring_00s_lm)$coefficients[1,1]
b.by.decade['Atlantic Herring', '2000s'] <- summary(herring_00s_lm)$coefficients[2,1]

herring_10s <- herring %>%
  filter(decade == '2010s')
herring_10s_lm <- lm(log(INDWT) ~ log(LENGTH), herring_10s)
a.by.decade['Atlantic Herring', '2010s'] <- summary(herring_10s_lm)$coefficients[1,1]
b.by.decade['Atlantic Herring', '2010s'] <- summary(herring_10s_lm)$coefficients[2,1]

# actual model!
herring_model <- lm(log(INDWT) ~ log(LENGTH)*decade, herring)
summary(herring_model)
```
<br>
Atlantic herring predictions plot, based on the model: the slopes of this relationship by decade appear to differ slightly. 

```{r herring predictions plot, echo=FALSE}
new_herring = expand.grid(LENGTH = seq(min(herring$LENGTH), max(herring$LENGTH), length.out = 100), decade = c("1990s", "2000s", "2010s")) 
new_herring$predicted <- exp(predict(herring_model, newdata = new_herring, type = "response")) 
new_herring$lower <- exp(predict(herring_model, newdata = new_herring, type = "response", interval = 
                                  "confidence")[,3])
new_herring$upper <- exp(predict(herring_model, newdata = new_herring, type = "response", interval = 
                                  "confidence")[,2])
ggplot(new_herring, aes(x = LENGTH, y = predicted))+
  geom_line(aes(color = decade))+
  geom_ribbon(aes(group = decade, ymax = upper, ymin = lower), alpha = 0.1)+ 
  scale_x_log10()+
  scale_y_log10()+ 
  theme_classic()+
  labs(title = "Atlantic Herring Predicted Weights by Length",
       x = "Length",
       y = "Individual Weight")
```

## Spiny Dogfish

Lastly, we analyze a fish following the equilibrium life strategy.

Original spiny dogfish data, length vs. weight: we see a clear exponential relationship between length and weight.

```{r original dogfish data plot, echo=FALSE}
dogfish <- lw_data %>%
  filter(COMNAME == 'SPINY DOGFISH') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2010 ~ "2010s",
                            YEAR >= 2000 & YEAR < 2010 ~ "2000s", 
                            YEAR < 2000 ~ "1990s"))

### ORIGINAL DATA PLOT
dogfish_orig <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point() +
  labs(title = "Spiny Dogfish: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(dogfish_orig)
```
<br>

After a log-transformation is applied, a clear linear pattern appears.

```{r transformed dogfish data plot, echo=FALSE}
### TRANSFORMED DATA PLOT
dogfish_log <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point() +
  labs(title = "Spiny Dogfish: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(dogfish_log)
```
<br>

Original data, color-coded by decade: we see the slope appears to decrease by decade again.

```{r original dogfish data by decade (color), echo=FALSE}
dogfish_dec.by.color.orig <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point(aes(color = decade))+
  labs(title = "Spiny Dogfish Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(dogfish_dec.by.color.orig)
```
<br>

Transformed data, color-coded by decade: the 2010s cluster appears more concentrated than the 1990s and 2000s clusters, and appears to have a lower maximum log(weight).

```{r transformed dogfish data by decade (color), echo=FALSE}
dogfish_dec.by.color.log <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point(aes(color = decade))+
  labs(title = "Spiny Dogfish: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(dogfish_dec.by.color.log)
```
<br>

Original data, split by decade: the slope of the relationship appears to decrease by decade.

```{r original dogfish data by decade (split), echo=FALSE}
### ORIGINAL DATA - SPLIT BY DECADE
dogfish_dec.split.orig <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point()+
  facet_wrap(~decade)+
  labs(title = "Spiny Dogfish: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(dogfish_dec.split.orig)
```
<br>

Transformed data, split by decade: this does not show as clear of a difference in the slope as the plot above.

```{r transformed dogfish data by decade (split), echo=FALSE}
dogfish_dec.split.log <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT)) +
  geom_point()+
  facet_wrap(~decade)+
  labs(title = "Spiny Dogfish: log(Length) vs log(Weight)",
       x = "log(Length)",
       y = "log(Individual Weight)")+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
print(dogfish_dec.split.log)
```
<br>

Spiny dogfish regression model: all terms are significant. Once again, the difference in the 2000s appears to be more significant than the 2010s.

```{r dogfish regression, echo=FALSE}
# actual statistics stuff once more! - saving parameters
dogfish_90s <- dogfish %>%
  filter(decade == '1990s')
dogfish_90s_lm <- lm(log(INDWT) ~ log(LENGTH), dogfish_90s)
a.by.decade['Spiny Dogfish', '1990s'] <- summary(dogfish_90s_lm)$coefficients[1,1]
b.by.decade['Spiny Dogfish', '1990s'] <- summary(dogfish_90s_lm)$coefficients[2,1]

dogfish_00s <- dogfish %>%
  filter(decade == '2000s')
dogfish_00s_lm <- lm(log(INDWT) ~ log(LENGTH), dogfish_00s)
a.by.decade['Spiny Dogfish', '2000s'] <- summary(dogfish_00s_lm)$coefficients[1,1]
b.by.decade['Spiny Dogfish', '2000s'] <- summary(dogfish_00s_lm)$coefficients[2,1]

dogfish_10s <- dogfish %>%
  filter(decade == '2010s')
dogfish_10s_lm <- lm(log(INDWT) ~ log(LENGTH), dogfish_10s)
a.by.decade['Spiny Dogfish', '2010s'] <- summary(dogfish_10s_lm)$coefficients[1,1]
b.by.decade['Spiny Dogfish', '2010s'] <- summary(dogfish_10s_lm)$coefficients[2,1]

# actual model!
dogfish_model <- lm(log(INDWT) ~ log(LENGTH)*decade, dogfish)
summary(dogfish_model)
```
<br>
Spiny dogfish predictions plot, based on the model: a slight difference in the slopes of the decadal lines can be seen here.

```{r dogfish predictions plot, echo=FALSE}
# generating predictions for the model
new_dogfish = expand.grid(LENGTH = seq(min(dogfish$LENGTH), max(dogfish$LENGTH), length.out = 100), decade = c("1990s", "2000s", "2010s")) 
new_dogfish$predicted <- exp(predict(dogfish_model, newdata = new_dogfish, type = "response")) 
new_dogfish$lower <- exp(predict(dogfish_model, newdata = new_dogfish, type = "response", interval = 
                                   "confidence")[,3])
new_dogfish$upper <- exp(predict(dogfish_model, newdata = new_dogfish, type = "response", interval = 
                                   "confidence")[,2])
ggplot(new_dogfish, aes(x = LENGTH, y = predicted))+
  geom_line(aes(color = decade))+
  geom_ribbon(aes(group = decade, ymax = upper, ymin = lower), alpha = 0.1)+ 
  scale_x_log10()+
  scale_y_log10()+ 
  theme_classic()+
  labs(title = "Spiny Dogfish Predicted Weights by Length",
       x = "Length",
       y = "Individual Weight")
```



## Parameters by decade

Slopes (b): 

```{r slopes table, echo=FALSE}
print(b.by.decade)
```
For cod, we see an increase in the slope from the 1990s to the 2000s and from the 1990s to the 2010s.

For American plaice, we seen an increase in the slope from the 1990s to the 2000s again, and a smaller increase from the 1990s to the 2010s.

In both cases, it increases from the 1990s to the 2000s and decreases slightly from the 2000s to the 2010s.

For Atlantic Herring, we see a similar pattern: an increase from the 1990s to the 2000s, but it appears that the smallest slope among the 3 decades appears in the 2010s this time. However, the difference from the 1990s to the 2010s is minimal.

The slope for spiny dogfish also peaks in the 2000s, and has its minimum in the 2010s.

These patterns make for an odd looking plot of the slopes over time. What happened in the 2010s?

```{r slopes plot, echo=FALSE}
b.to.df <- data.frame(
  Species = c("Cod", "American Plaice", "Atlantic Herring", "Spiny Dogfish"),
  `1990s` = c(3.075384, 3.281383, 3.048610, 3.124006),
  `2000s` = c(3.104295, 3.341870, 3.126295, 3.179436),
  `2010s` = c(3.080731, 3.305686, 3.040111, 3.100683),
  stringsAsFactors = FALSE
)
b.long <- pivot_longer(b.to.df, 
                        cols = starts_with("X"), 
                        names_to = "Decade", 
                        values_to = "Slope")
ggplot(b.long, aes(x = Decade, y = Slope, group = Species, color = Species)) +
  geom_line() +
  geom_point() +
  labs(title = "Slope of Relationship Between log(Length) and log(Weight) By Decade", x = "Decade", y = "Slope (b)") +
  theme_classic()
```



Intercepts (a):

```{r intercepts table, echo=FALSE}
print(a.by.decade)
```
A few directions to go from here - I think checking differences in the slope based on location would be most interesting (Northern Gulf of Maine vs. Southern Gulf of Maine, maybe). My hypothesis is that the fish that migrated farther north to avoid warming water (the "effective trackers," per Kathy's paper), may be biasing results towards the null as their sizes likely haven't changed if they maintained a constant temperature. Even if this logic is completely wrong, I think that might be cool to look at.

I also might need some help on handling the variance with the log-log transformation/understanding what handling it means in terms of the analysis - I think some the math in that .html went over my head a bit. 