---
title: "Statistics Week 5"
output: html_document
date: "2024-07-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
<br>
The models in this file are based off of the years being divided up in 10(ish) year increments: 1992-2002, 2003-2013, and 2014-2023. 

```{r setup 2, include=FALSE}
library(tidyverse) # loading in packages & data
library(emmeans)
lw_data <- readRDS('Data/length_weight_epu.rds')
lw_coeffs <- readRDS('Data/lw_coeffs.rds')

# Table to hold b parameter
b.by.10yrs <- matrix(nrow=4, ncol=3, byrow=TRUE)
rownames(b.by.10yrs) <- c('Cod', 'American Plaice', 'Atlantic Herring', 'Spiny Dogfish')
colnames(b.by.10yrs) <- c('1992 - 2002', '2003 - 2013', '2014 - 2023')

# Table to hold log(a) parameter
a.by.10yrs <- matrix(nrow=4, ncol=3, byrow=TRUE)
rownames(a.by.10yrs) <- c('Cod', 'American Plaice', 'Atlantic Herring', 'Spiny Dogfish')
colnames(a.by.10yrs) <- c('1992 - 2002', '2003 - 2013', '2014 - 2023')
```

## Atlantic Cod

```{r cod plot}
cod <- lw_data %>%
  filter(COMNAME == 'ATLANTIC COD') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2014 ~ "2014 - 2023",
                            YEAR >= 2003 & YEAR < 2014 ~ "2003 - 2013",
                            YEAR <= 2002 ~ "1992 - 2002")) %>%
  filter(!is.na(EPU))

### Cod Data Plot
cod_plot <- cod %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=decade)) +
  facet_wrap(~EPU)+
  labs(title = "Atlantic Cod: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(cod_plot)
```
<br>
```{r cod model}
cod_model <- lm(log(INDWT) ~ log(LENGTH)*decade + log(LENGTH)*EPU, data=cod)
summary(cod_model)
a.by.10yrs["Cod", "1992 - 2002"] <- cod_model$coefficients[1] + cod_model$coefficients[5]
b.by.10yrs["Cod", "1992 - 2002"] <- cod_model$coefficients[2] + cod_model$coefficients[10]
a.by.10yrs["Cod", "2003 - 2013"] <- cod_model$coefficients[1] + cod_model$coefficients[5] + cod_model$coefficients[3]
b.by.10yrs["Cod", "2003 - 2013"] <- cod_model$coefficients[2] + cod_model$coefficients[10] + cod_model$coefficients[8]
a.by.10yrs["Cod", "2014 - 2023"] <- cod_model$coefficients[1] + cod_model$coefficients[5] + cod_model$coefficients[4]
b.by.10yrs["Cod", "2014 - 2023"] <- cod_model$coefficients[2] + cod_model$coefficients[10] + cod_model$coefficients[9]
```
<br>
Cod predictions plots: one is faceted by decade and one by EPU.
```{r cod predictions plot}
# Cod Predictions Plot - facet by EPU
cod$decade <- factor(cod$decade)
cod$EPU <- factor(cod$EPU)
new_cod <- expand.grid(LENGTH = seq(min(cod$LENGTH), max(cod$LENGTH), length.out = 100), 
  decade = levels(cod$decade),
  EPU = levels(cod$EPU))
new_cod$decade <- factor(new_cod$decade, levels = levels(cod$decade))
new_cod$EPU <- factor(new_cod$EPU, levels = levels(cod$EPU))
predictions <- predict(cod_model, newdata = new_cod, interval = "confidence")
new_cod$predicted <- exp(predictions[, "fit"])
new_cod$lower <- exp(predictions[, "lwr"])
new_cod$upper <- exp(predictions[, "upr"])

ggplot(new_cod, aes(x = LENGTH, y = predicted, color = decade)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = decade), alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  theme_light() +
  labs(title = "Cod Predicted Weights by Length",
      x = "Length",
      y = "Individual Weight") +
  facet_wrap(~EPU)

# Cod Predictions Plot - facet by decade
cod$decade <- factor(cod$decade)
cod$EPU <- factor(cod$EPU)
new_cod <- expand.grid(LENGTH = seq(min(cod$LENGTH), max(cod$LENGTH), length.out = 100), 
  decade = levels(cod$decade),
  EPU = levels(cod$EPU))

predictions <- predict(cod_model, newdata = new_cod, interval = "confidence")

new_cod$predicted <- exp(predictions[, "fit"])
new_cod$lower <- exp(predictions[, "lwr"])
new_cod$upper <- exp(predictions[, "upr"])

ggplot(new_cod, aes(x = LENGTH, y = predicted, color = EPU)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = EPU), alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  theme_light() +
  labs(
    title = "Cod Predicted Weights by Length",
    x = "Length",
    y = "Individual Weight"
  ) +
  facet_wrap(~decade)
```
<br>
Cod weight comparisons at the mean length:
```{r cod comparisons}
cod.conts.epu <- emmeans(cod_model, ~EPU, by = "LENGTH", data=cod)
pairs(cod.conts.epu)

cod.conts.dec <- emmeans(cod_model, ~decade, by = "LENGTH", data=cod)
pairs(cod.conts.dec)
```

## American Plaice

```{r plaice plot}
plaice <- lw_data %>%
  filter(COMNAME == 'AMERICAN PLAICE') %>%
  filter(INDWT != 0 & INDWT < 4) %>%
  mutate(decade = case_when(YEAR >= 2014 ~ "2014 - 2023",
                            YEAR >= 2003 & YEAR < 2014 ~ "2003 - 2013",
                            YEAR <= 2002 ~ "1992 - 2002")) %>%
  filter(!is.na(EPU))

### American Plaice Data Plot
plaice_plot <- plaice %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=decade)) +
  facet_wrap(~EPU)+
  labs(title = "American Plaice: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(plaice_plot)
```
<br>
```{r plaice model}
plaice_model <- lm(log(INDWT) ~ log(LENGTH)*decade + log(LENGTH)*EPU, data=plaice)
summary(plaice_model)
a.by.10yrs["American Plaice", "1992 - 2002"] <- plaice_model$coefficients[1] + plaice_model$coefficients[5]
b.by.10yrs["American Plaice", "1992 - 2002"] <- plaice_model$coefficients[2] + plaice_model$coefficients[10]
a.by.10yrs["American Plaice", "2003 - 2013"] <- plaice_model$coefficients[1] + plaice_model$coefficients[5] + plaice_model$coefficients[3]
b.by.10yrs["American Plaice", "2003 - 2013"] <- plaice_model$coefficients[2] + plaice_model$coefficients[10] + plaice_model$coefficients[8]
a.by.10yrs["American Plaice", "2014 - 2023"] <- plaice_model$coefficients[1] + plaice_model$coefficients[5] + plaice_model$coefficients[4]
b.by.10yrs["American Plaice", "2014 - 2023"] <- plaice_model$coefficients[2] + plaice_model$coefficients[10] + plaice_model$coefficients[9]
```
<br>
American plaice predictions plots: one is faceted by decade and one by EPU.
```{r plaice predictions plot}
# American Plaice Predictions Plot - facet by EPU
plaice$decade <- factor(plaice$decade)
plaice$EPU <- factor(plaice$EPU)
new_plaice <- expand.grid(LENGTH = seq(min(plaice$LENGTH), max(plaice$LENGTH), length.out = 100), 
  decade = levels(plaice$decade),
  EPU = levels(plaice$EPU))

predictions <- predict(plaice_model, newdata = new_plaice, interval = "confidence")

new_plaice$predicted <- exp(predictions[, "fit"])
new_plaice$lower <- exp(predictions[, "lwr"])
new_plaice$upper <- exp(predictions[, "upr"])

ggplot(new_plaice, aes(x = LENGTH, y = predicted, color = decade)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = decade), alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  theme_light() +
  labs(
    title = "American Plaice Predicted Weights by Length",
    x = "Length",
    y = "Individual Weight"
  ) +
  facet_wrap(~EPU)

# American Plaice Predictions Plot - facet by decade
plaice$decade <- factor(plaice$decade)
plaice$EPU <- factor(plaice$EPU)
new_plaice <- expand.grid(LENGTH = seq(min(plaice$LENGTH), max(plaice$LENGTH), length.out = 100), 
  decade = levels(plaice$decade),
  EPU = levels(plaice$EPU))

predictions <- predict(plaice_model, newdata = new_plaice, interval = "confidence")

new_plaice$predicted <- exp(predictions[, "fit"])
new_plaice$lower <- exp(predictions[, "lwr"])
new_plaice$upper <- exp(predictions[, "upr"])

ggplot(new_plaice, aes(x = LENGTH, y = predicted, color = EPU)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = EPU), alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  theme_light() +
  labs(
    title = "American Plaice Predicted Weights by Length",
    x = "Length",
    y = "Individual Weight"
  ) +
  facet_wrap(~decade)
```
<br>
American plaice weight comparisons at the mean length:
```{r plaice comparisons}
plaice.conts.epu <- emmeans(plaice_model, ~EPU, by = "LENGTH", data=plaice)
pairs(plaice.conts.epu)

plaice.conts.dec <- emmeans(plaice_model, ~decade, by = "LENGTH", data=plaice)
pairs(plaice.conts.dec)
```

## Atlantic Herring

```{r herring plot}
herring <- lw_data %>%
  filter(COMNAME == 'ATLANTIC HERRING') %>%
  filter(INDWT != 0 & INDWT < 4) %>%
  mutate(decade = case_when(YEAR >= 2014 ~ "2014 - 2023",
                            YEAR >= 2003 & YEAR < 2014 ~ "2003 - 2013",
                            YEAR <= 2002 ~ "1992 - 2002")) %>%
  filter(!is.na(EPU))

### Atlantic Herring Data Plot
herring_plot <- herring %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=decade)) +
  facet_wrap(~EPU)+
  labs(title = "Atlantic Herring: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(herring_plot)
```
<br>
```{r herring model}
herring_model <- lm(log(INDWT) ~ log(LENGTH)*decade + log(LENGTH)*EPU, data=herring)
summary(herring_model)
a.by.10yrs["Atlantic Herring", "1992 - 2002"] <- herring_model$coefficients[1] + herring_model$coefficients[5]
b.by.10yrs["Atlantic Herring", "1992 - 2002"] <- herring_model$coefficients[2] + herring_model$coefficients[10]
a.by.10yrs["Atlantic Herring", "2003 - 2013"] <- herring_model$coefficients[1] + herring_model$coefficients[5] + herring_model$coefficients[3]
b.by.10yrs["Atlantic Herring", "2003 - 2013"] <- herring_model$coefficients[2] + herring_model$coefficients[10] + herring_model$coefficients[8]
a.by.10yrs["Atlantic Herring", "2014 - 2023"] <- herring_model$coefficients[1] + herring_model$coefficients[5] + herring_model$coefficients[4]
b.by.10yrs["Atlantic Herring", "2014 - 2023"] <- herring_model$coefficients[2] + herring_model$coefficients[10] + herring_model$coefficients[9]
```
<br>
Atlantic herring predictions plots: one is faceted by decade and one by EPU.
```{r herring predictions plot}
# Atlantic Herring Predictions Plot - facet by EPU
herring$decade <- factor(herring$decade)
herring$EPU <- factor(herring$EPU)
new_herring <- expand.grid(LENGTH = seq(min(herring$LENGTH), max(herring$LENGTH), length.out = 100), decade = levels(herring$decade), EPU = levels(herring$EPU))

predictions <- predict(herring_model, newdata = new_herring, interval = "confidence")

new_herring$predicted <- exp(predictions[, "fit"])
new_herring$lower <- exp(predictions[, "lwr"])
new_herring$upper <- exp(predictions[, "upr"])

ggplot(new_herring, aes(x = LENGTH, y = predicted, color = decade)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = decade), alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  theme_light() +
  labs(
    title = "Atlantic Herring Predicted Weights by Length",
    x = "Length",
    y = "Individual Weight"
  ) +
  facet_wrap(~EPU)

# Atlantic Herring Predictions Plot - facet by decade
herring$decade <- factor(herring$decade)
herring$EPU <- factor(herring$EPU)
new_herring <- expand.grid(LENGTH = seq(min(herring$LENGTH), max(herring$LENGTH), length.out = 100), 
  decade = levels(herring$decade),
  EPU = levels(herring$EPU))

predictions <- predict(herring_model, newdata = new_herring, interval = "confidence")

new_herring$predicted <- exp(predictions[, "fit"])
new_herring$lower <- exp(predictions[, "lwr"])
new_herring$upper <- exp(predictions[, "upr"])

ggplot(new_herring, aes(x = LENGTH, y = predicted, color = EPU)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = EPU), alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  theme_light() +
  labs(
    title = "Atlantic Herring Predicted Weights by Length",
    x = "Length",
    y = "Individual Weight"
  ) +
  facet_wrap(~decade)
```
<br>
Atlantic herring weight comparisons at the mean length:
```{r herring comparisons}
herring.conts.epu <- emmeans(herring_model, ~EPU, by = "LENGTH", data=herring)
pairs(herring.conts.epu)

herring.conts.dec <- emmeans(herring_model, ~decade, by = "LENGTH", data=herring)
pairs(herring.conts.dec)
```

## Spiny Dogfish

```{r dogfish plot}
dogfish <- lw_data %>%
  filter(COMNAME == 'SPINY DOGFISH') %>%
  filter(INDWT != 0) %>%
  mutate(decade = case_when(YEAR >= 2014 ~ "2014 - 2023",
                            YEAR >= 2003 & YEAR < 2014 ~ "2003 - 2013",
                            YEAR <= 2002 ~ "1992 - 2002")) %>%
  filter(!is.na(EPU))

### Spiny dogfish Data Plot
dogfish_plot <- dogfish %>%
  ggplot(aes(x = LENGTH, y = INDWT))+
  geom_point(aes(color=decade)) +
  facet_wrap(~EPU)+
  labs(title = "Spiny Dogfish: Length vs Weight",
       x = "Length",
       y = "Individual Weight")+
  theme_bw()
print(dogfish_plot)
```
<br>
```{r dogfish model}
dogfish_model <- lm(log(INDWT) ~ log(LENGTH)*decade + log(LENGTH)*EPU, data=dogfish)
summary(dogfish_model)
a.by.10yrs["Spiny Dogfish", "1992 - 2002"] <- dogfish_model$coefficients[1] + dogfish_model$coefficients[5]
b.by.10yrs["Spiny Dogfish", "1992 - 2002"] <- dogfish_model$coefficients[2] + dogfish_model$coefficients[10]
a.by.10yrs["Spiny Dogfish", "2003 - 2013"] <- dogfish_model$coefficients[1] + dogfish_model$coefficients[5] + dogfish_model$coefficients[3]
b.by.10yrs["Spiny Dogfish", "2003 - 2013"] <- dogfish_model$coefficients[2] + dogfish_model$coefficients[10] + dogfish_model$coefficients[8]
a.by.10yrs["Spiny Dogfish", "2014 - 2023"] <- dogfish_model$coefficients[1] + dogfish_model$coefficients[5] + dogfish_model$coefficients[4]
b.by.10yrs["Spiny Dogfish", "2014 - 2023"] <- dogfish_model$coefficients[2] + dogfish_model$coefficients[10] + dogfish_model$coefficients[9]
```
<br>
Spiny Dogfish predictions plots: one is faceted by decade and one by EPU.
```{r dogfish predictions plot}
# Spiny dogfish Predictions Plot - facet by EPU
dogfish$decade <- factor(dogfish$decade)
dogfish$EPU <- factor(dogfish$EPU)
new_dogfish <- expand.grid(LENGTH = seq(min(dogfish$LENGTH), max(dogfish$LENGTH), length.out = 100), 
  decade = levels(dogfish$decade),
  EPU = levels(dogfish$EPU))

predictions <- predict(dogfish_model, newdata = new_dogfish, interval = "confidence")

new_dogfish$predicted <- exp(predictions[, "fit"])
new_dogfish$lower <- exp(predictions[, "lwr"])
new_dogfish$upper <- exp(predictions[, "upr"])

ggplot(new_dogfish, aes(x = LENGTH, y = predicted, color = decade)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = decade), alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  theme_light() +
  labs(
    title = "Spiny Dogfish Predicted Weights by Length",
    x = "Length",
    y = "Individual Weight"
  ) +
  facet_wrap(~EPU)

# Spiny Dogfish Predictions Plot - facet by decade
dogfish$decade <- factor(dogfish$decade)
dogfish$EPU <- factor(dogfish$EPU)
new_dogfish <- expand.grid(LENGTH = seq(min(dogfish$LENGTH), max(dogfish$LENGTH), length.out = 100), 
  decade = levels(dogfish$decade),
  EPU = levels(dogfish$EPU))

predictions <- predict(dogfish_model, newdata = new_dogfish, interval = "confidence")

new_dogfish$predicted <- exp(predictions[, "fit"])
new_dogfish$lower <- exp(predictions[, "lwr"])
new_dogfish$upper <- exp(predictions[, "upr"])

ggplot(new_dogfish, aes(x = LENGTH, y = predicted, color = EPU)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = EPU), alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  theme_light() +
  labs(
    title = "Spiny Dogfish Predicted Weights by Length",
    x = "Length",
    y = "Individual Weight"
  ) +
  facet_wrap(~decade)
```
<br>
Spiny dogfish weight comparisons at the mean length:
```{r dogfish comparisons}
dogfish.conts.epu <- emmeans(dogfish_model, ~EPU, by = "LENGTH", data=dogfish)
pairs(dogfish.conts.epu)

dogfish.conts.dec <- emmeans(dogfish_model, ~decade, by = "LENGTH", data=dogfish)
pairs(dogfish.conts.dec)
```

## Slope Changes for Gulf of Maine Fish

```{r slope changes}
b.by.10yrs.df <- as.data.frame(b.by.10yrs, stringsAsFactors = FALSE)
b.by.10yrs.df$Species <- rownames(b.by.10yrs)

b.by.10yrs.df.long <- pivot_longer(b.by.10yrs.df, 
                            cols = c("1992 - 2002", "2003 - 2013", "2014 - 2023"), 
                            names_to = "Decade", 
                            values_to = "Slope")

ggplot(b.by.10yrs.df.long, aes(x = Decade, y = Slope, group = Species, color = Species)) +
  geom_line() +
  geom_point() +
  labs(title = "Slope of Relationship Between log(Length) and log(Weight) in the Gulf of Maine", 
       x = "Decade", 
       y = "Slope (b)") +
  theme_minimal()
```

## All Species Coefficients by Season

On par with Wigley's work:

```{r coeffs, message=FALSE}
unique_comname <- unique(lw_data$COMNAME)
unique_comname <- na.omit(unique_comname)

data <- lw_data %>%
  filter(!is.na(LENGTH) & !is.na(INDWT) & !is.na(COMNAME)) %>%
  mutate(decade = case_when(YEAR >= 2014 ~ "2014 - 2023",
                            YEAR >= 2003 & YEAR < 2014 ~ "2003 - 2013",
                            YEAR <= 2002 ~ "1992 - 2002"))

grouped_data <- data %>%
  group_by(COMNAME, decade) %>%
  summarise(count = n())

filtered_groups <- grouped_data %>%
  filter(count >= 5) %>%
  ungroup()

unique_species <- unique(filtered_groups$COMNAME)
species <- c()

for (spec in unique_species) {
  spec_decs <- filtered_groups %>%
    filter(COMNAME == spec)
  if (nrow(spec_decs) == 3) {
    species <- c(species, spec)
  }
}

season_counts <- data %>%
  filter(COMNAME %in% species) %>% 
  filter(!is.na(SEASON)) %>%             
  group_by(COMNAME, decade, SEASON) %>%
  summarise(count = n()) %>%
  filter(count >= 5) %>%
  ungroup()

####

coeffs.by.decade.df <- data.frame(row.names = species)
decades = c('1992 - 2002', '2003 - 2013', '2014 - 2023')
seasons = c('SPRING', 'FALL')

# filtering out large outliers
data <- data %>%
  filter(! (COMNAME == "AMERICAN PLAICE" & INDWT > 4) ) %>%
  filter(! (COMNAME == "ATLANTIC HERRING" & INDWT > 0.5)) %>%
  filter(! (COMNAME == "ATLANTIC SHARPNOSE SHARK" & LENGTH > 106)) %>%
  ## bluntnose stingray - says they can be up to 100 cm. a few observations are close/over
  filter(! (COMNAME == "BLUNTNOSE STINGRAY" & INDWT > 50)) %>%
  filter(! (COMNAME == "BLUNTNOSE STINGRAY" & INDWT > 10 & LENGTH < 10)) %>%
  # just some low observations... not sure if these are errors
  filter(! (COMNAME == "BUTTERFISH" & INDWT <= 0.01 & LENGTH > 12.5)) %>%
  filter(! (COMNAME == "LITTLE SKATE" & INDWT > 4)) %>%
  filter(! (COMNAME == "LONGFIN SQUID" & INDWT > 1)) %>%
  filter(! (COMNAME == "LONGHORN SCULPIN" & INDWT > 750)) %>%
  filter(! (COMNAME == "NORTHERN KINGFISH" & INDWT > 1)) %>%
  filter(! (COMNAME == "NORTHERN SHORTFIN SQUID" & INDWT > 1)) %>%
  filter(! (COMNAME == "WINTER FLOUNDER" & LENGTH < 20 & INDWT > 3))

for (d in decades) {
  for (s in seasons) {
    coeffs.by.decade.df[[paste0("B.", s, ".", d)]] <- NA
    coeffs.by.decade.df[[paste0("A.", s, ".", d)]] <- NA
  }
}

for (spec in species) {
  spec_data <- lw_data %>%
    filter(COMNAME == spec) %>%
    filter(INDWT != 0) %>%
    mutate(decade = case_when(YEAR >= 2014 ~ "2014 - 2023",
                            YEAR >= 2003 & YEAR < 2014 ~ "2003 - 2013",
                            YEAR <= 2002 ~ "1992 - 2002"))
  
  for (dec in decades) {
    for (szn in seasons) {
      season_data <- spec_data %>%
        filter(SEASON == szn & decade == dec)
      if (nrow(season_data) >= 5) {
        spec.szn.lm <- lm(log(INDWT) ~ log(LENGTH), data = season_data)
        coeffs.by.decade.df[spec, paste0("B.", szn, ".", dec)] <- coef(spec.szn.lm)[2]
        coeffs.by.decade.df[spec, paste0("A.", szn, ".", dec)] <- coef(spec.szn.lm)[1]
      }
    }
  }
}

print(coeffs.by.decade.df)
```

